The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region

### 2017-02-26 paper.tex
Add latex paper

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação Getulio Vargas) and Robin Lovelace (University of Leeds)},
            pdftitle={The unequal distribution of access to education and transport: insights from for São Paulo},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{The unequal distribution of access to education and transport: insights
from for São Paulo}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos
(Fundação Getulio Vargas) and Robin Lovelace (University of Leeds)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

@@

In many large Latin American cities such as São Paulo, growing social
and economic inequalities are embedded through unfair education and
transport systems. Good schools are mostly concentrated in wealthy
areas, while transport links to school are deficient in deprived areas,
exacerbating the issue. Inequalities in educational and transport
infrastructure are mutually reinforcing: the right to mobility is
intrinsically linked to the right to education. This is manifested by
the overlap between recent protests against unwanted changes to public
education and the social movements contesting increases in public
transport fares. Another manifestation is to be found in the concept of
school accessibility. This paper sheds light on the transport-education
inequality nexus with reference to a new school accessibility measure
applied São Paulo. By capturing both the unequal distribution of schools
and transport services across space, the index allows embedded
inequalities to be better understood and, with political will,
contested. Our index combines information on the spatial distribution of
children and adolescents by income categories, the location of existing
schools, the travel patterns of students, and the travel infrastructure
serving the school catchment area into a single measure. The index is
used to measure school accessibility locally across São Paulo, using
data sources from Population and School Censuses, commonly available in
Latin American cities. The results illustrate how existing inequalities
are amplified by variable accessibility to schools across income groups
and geographical space. We conclude that extending the concept of local
accessibility indicators to education can help to both contest and
constructively tackle embedded social inequalities.

@@

\textbf{Contents:}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Motivations
\item
  General measures of accessibility
\item
  School accessibility
\end{itemize}

\ldots{}

Travel to school is an everyday reality for millions of young people
around the world. The mode, duration, safety, comfort and pollution
levels of this trip has huge impacts on the future generation, yet has
received relatively little academic attention.

Travel options are vital for ensuring a more equitable supply of
educational opportunity to diverse groups. Conversely, poor
accessibility to deprived area can reinforce social inequalities, with
long-term implications. Based on this emerging context, this paper
develops a school accessibility index for local areas.

The first well-known attempt to define accessibility quantitatively was
by Ingram (1971), which presented a range of measures related to
distance (Euclidean and network), barriers and different functions
representing distance decay.

This early work made the distinction between accessibility indeces that
apply to zones or single `desire lines': ``relative accessibility is
defined as a measure of the effort of overcoming spatial separation
between two points, while the integral accessibility is defined as a
measure of the effort of overcoming spatial separation between a point
and all other points within an area'' (Allen, Liu, and Singer 1993).

\section{Method: measuring school
accessibility}\label{method-measuring-school-accessibility}

We calculate three different accessibility measures for different modes:
cumulative, gravitiy-based, and competitive.

The cumulative-opportunity measure for mode \(M\) (Boisjoly and
El-Geneidy, 2016) is defined as:

\[ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) \]

\[ f(C_{ij}) = \left\{ 
                \begin{array}{ll}
                  1 if C_{ij}<=t\\
                  0 if C_{ij}<t
                \end{array}
              \right.
  \]

This measure counts the number of schools available from one area within
a certain travel time by mode \(M\) threshold. \(C_{ij}\) is the travel
cost (measured in time) between the centroid of zone i and the centroid
of zone j.

Gravity-based accessibility measure

In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found
that the cumulative-opportunity measure (using a 45 minute threshold) is
highly correlated with a gravity-based accessibility measure
(({\textbf{???}}), ({\textbf{???}})), defined as:

\[ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) \]

Where \(A_{i}^M\) is the gravity accessibility at the centroid of area
\(i\) to all schools at area \(j\) using mode \(M\), and \(\beta\) is a
negative exponential cost function. The cost function, based on a
negative exponential decay curve, includes reported number of school
trips in the 2007 Origin-Destination Survey. (Here we can also use
information from the school census on number of students per school to
know how many students the area attracts). The proposed accessibility
measure captures the fact that more proximate schools (S) are more
attractive than those located further away.

Competitive accessibility measure

The measures proposed above take into account the sptail distribution of
schooling opportunities, but not the local demand for schooling. This is
particularly relevant for our equity analysis, since it could be the
case that higher areas are disproportionally served with respect to the
number of potentail students living within a certain travel distance,
whereas the opposite holds for lower-income areas. In order to assess
the mismatch between the demand and supply for schooling, we use the sum
of students in schools in each area (the demand), and the sum of
individuals within the school grade age-group living in each area (the
supply) in the following competitive accessibility measure, first
proposed by ({\textbf{???}}) and adapted by El-Geneidy et al (2015):

\[ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}\]

Where the numerator discounts the number of pupils in area \(j\)
\(P_{j}\) by how far area \(i\) is from area \(j\) using the same
function as before, and the denominator discounts the number of students
(young adults) living in zone \(k\) \(Y_{k}\) by how far they are from
area \(j\). In this way, the discounted number of study places at each
area is divided by the potential students available to fill those
places, and then summed in order to obtain a single accessibility
measure for each area \(i\).

In order to assess equality in access to schooling, we stratify our
population of students in two student groups (we could do it for more
income groups): the students living with a head of household who earns 1
minimum wage or less, and those living with a head of household who
earns more than 1 minimum wage. The accessibility measures are based on
student's travel times by different modes using data from the 2007
Origin-Destination survey. Alternatively, for the case of public
transport, wet use actual transit times in 2015 for comparison (bilhete
unico data?).

\section{Results}\label{results}

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Descriptive statistics about modal choice and travel time by AEPs and
  level of income
\item
  Spatial Distribution of Schools and Transit-Dependent Students at the
  AEP Level
\end{itemize}

\section{Discussion}\label{discussion}

\section{Conclusion}\label{conclusion}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)},
            pdftitle={The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

\setlength{\droptitle}{-2em}
  \title{The unequal distribution of access to education and transport: insights
from the São Paulo Metropolitan Region}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Ana I. Moreno-Monroy (Autonomous university of Barcelona), Robin Lovelace
(University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \date{}
  \predate{}\postdate{}

@@

In many large Latin American urban areas such as the São Paulo
Metropolitan Region (SPMR), growing social and economic inequalities are
embedded through high spatial inequality in the provision of education
and transport systems. This paper sheds light on the transport-education
inequality nexus with reference to a new school accessibility measure
applied to the SPMR. Our index combines information on the spatial
distribution of adolescents, the location of existing schools, and the
travel infrastructure serving the school catchment area into a single
measure. The index is used to measure school accessibility locally
across 633 areas within the SPMR. We use the index to simulate the
impact of a policiy aiming at increasing the centralization of public
secondary education provision, and find that it negatively affects
students with the lowest levels of accessibility. These results
illustrate how existing inequalities are amplified by variable
accessibility to schools across income groups and geographical space.

\bigskip

\textit{This version: January 2017. Preliminary version. Please do not cite}

@@

Inequalities in educational and transport infrastructure are mutually
reinforcing: the right to mobility is intrinsically linked to the right
to education. Travel to school options are vital for ensuring a more
equitable supply of educational opportunity to diverse groups.
Conversely, poor accessibility to and from deprived areas can reinforce
social inequalities, with long-term implications. Against this
background, in this paper we propose a new way to measure school
accessibility in local areas within an urban area, and apply it for the
case of the São Paulo Metropolitan Region (SPMR).

The SPMR presents an excellent opportunity to measure the extent of
educational and transport infrastructure, and to study the impact of
public policies on different socio-economic groups. Partly as a response
to the pressures from the Free Pass Movement that started in March,
2013, low income public school students residing in the SPMR gained
access to a public transport subsidy in 2015. However, in October 2015,
the São Paulo state government announced during a television interview
that as part of a budgetary deficit reduction plan, dozens of secondary
schools would be closed in 2016. The change in policy was estimated to
affect more than 300,000 students, many whom would be placed in schools
far from their homes. As a response, in the following months students
occupied over 200 schools. Protesters scored some victories: the reform
was postponed for one year, and the education secretary resigned
(Ortellado 2015).

Although this time the policy change was not carried out, in a context
of high spatial inequalities in the provision of public transport and
schooling, it is worth asking whether a public transport subsidy for
students can compensate for the lower provision of public schools in
some areas. In this dynamic and highly political context, new
quantitative evidence can help shed light on the relationship between
transport and educational inequalities and the extent to which they are
mutually reinforcing. Moreover, the results should help design policies
that are more inclusive and equitable by highlighting areas where the
provision of schools, travel to school opportunities, or both, are
particularly restrictive.

Our proposed competitive accessibility measure combines information
about the place of residence of students, the spatial distribution of
public schools, and public transport accessibility into a single
measure. We start from a cumulative opportunity measure that counts the
number of schools seats that can be reached within a 30 minute journey
by public transport, and then build a competitive measure that takes
into account the place of residency of teens. Travel-to-school times by
public transport are based on actual commuting times obtained through
the Google Matrix Distance API. We calculate these measures for 633
areas within the SPMR. We use the competitive measure to simulate the
impact of a policy change in the location of public secondary schools,
in order to, first, understand the extent of spatial inequalities in
school accessibility by public transport, and second, estimate the
effect of a policy aiming at concentrating public secondary education
provision on school accessibility. We find that closing down schools in
areas with lower than average provision is highly regressive: it
negatively impacts students with the lowest accessibility levels,
accentuating existing inequalities.

The first well-known attempt to define accessibility quantitatively was
by Ingram (1971), which presented a range of measures related to
distance (Euclidean and network), barriers and different functions
representing distance decay. This early work made the distinction
between accessibility indices that apply to zones or single `desire
lines': ``relative accessibility is defined as a measure of the effort
of overcoming spatial separation between two points, while the integral
accessibility is defined as a measure of the effort of overcoming
spatial separation between a point and all other points within an area''
(Allen, Liu, and Singer 1993). In subsequent works, the measurement of
job accessibility by particular transport modes has received vast
attention in the literature (Geurs and Van Wee 2004), while studies on
access to education has largely focused on the consequences of mode
choice on socio-economic indicators and school outcomes (see for
instance Asahi (2014), Falch, Lujala, and Str{ø}m (2013) and Dickerson
and McIntosh (2012) on the impact of accessibility on school outcomes,
and Andersson, Malmberg, and {Ö}sth (2012) and Easton and Ferrari (2015)
on the effect of changes in travel-to-school patterns in developed
economies).

Our index is inspired by the index developed by Shen (1998) in the
context of job accessibility. The main innovative element with respect
to previous proposed measures of school accessibility is that it takes
into account both the `supply' and `demand' for schooling in each area.
Concretely, it is based on the idea that in each local area, there is a
certain amount of students in (secondary) school age, and a certain
quantity of (public) school seats available. As areas are part of a
city, each local area is also subject to the inflow of potential
students from other areas, as well of the outflow of students to other
areas. The magnitude of the net flow will depend on the travel distance
between all other areas and the area in question, which here we consider
to be the public transport commuting time between areas. The index thus
captures the fact that students compete for school seats, which are
limited in number, and distributed unequally across space. The inclusion
of this competitive element to school accessibility, akin to that in job
access, highlights the fact that under certain educational systems, the
access to opportunities is mediated by competition.

Our work contributes to the literature on transport-related social
exclusion (Jones and Lucas 2012) (Lucas et al. 2016) by providing a
quantitative way to assess multiple dimensions into a single measure.
Education and public transport are both essential public services, and
thus are part of the set of fundamental and universal human rights. Lack
of sufficient access to these services limits the way in which
individuals use can their capacities and exersise their rights in a
context of equal opportunities (Gomide 2003). Our index provides a
benchmark to compare access inequalities in countries with unequal
provision of public services (Gomide 2006), (Pacione 1989). Considering
contestation as the process through which social groups mobilize in an
organized way in order to impede the implementation of unwanted
policies, or force the negotiation of new conditions, our index allows
embedded inequalities to be better understood.

After this introduction, we detail our data sources and definitions in
Section 2. Section 3 describes the area of study and presents some
preliminary findings. Section 4 describes the proposed methods for
measuring school accessibility. Section 5 presents the results for the
SPMR, and finally, Section 6 concludes.

\section{2. Data}\label{data}

z school accessibility index, we use data from three
different sources. The first one is the 2008 School Census of the
Brazilian National Institute of Educational Studies and Surveys (INEP),
which provides information on the (universe) of public educational
institutions, including a unique identifier and the number of students
enrolled in secondary education. We use the coordinates contained in the
database provided by the Center for Metropolitan Studies CEM
(\textit{Centro de Estudos da Metrópole}) for 2001 to geolocate the
schools in 2008 using the unique identifier. In order to include new
schools built between 2001 and 2008, we used a dataset provided by the
Fundação SEADE (São Paulo State Agency for Statistical Analysis)
containing the goecoded location for all state administration schools at
2008. The final dataset set was obtained merging this two data source
using the standard ID used by the INEP. We geo-localized a total of
4,612 public schools in the SPRM.

Our second source is information the 2010 Population Census of Brazil,
compiled and freely distributed by the Brazilian Institute of Statistics
(IBGE). We aggregate the data by Area de Ponderacão Espacial (AEP)
areas, a spatial unit defined for surveying purposes by IBGE. There is a
total of 633 areas in the SPMR. IBGE provides the digital networks
containing the boundaries of these areas. From the census microdata, we
have data on the number of inhabitants by age for each enumeration area.
The number of adolescents in secondary school age (ages 15-18) is
1'216.611 (approximately 13 percent of the population).

The final source of information was derived from the Google Distance
Matrix API, which provided travel times and distances by public
transport. This was implemented using an interface to the API by means
of the R package \textbf{stplanr} (Lovelace et al. 2016). To provide an
example of how the method works, a reproducible example is illustrated
below.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(stplanr)}
\NormalTok{o =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.31813}\NormalTok{, -}\FloatTok{23.38584}\NormalTok{) }\CommentTok{# coordinates of origin}
\NormalTok{d =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.34078}\NormalTok{, -}\FloatTok{23.37746}\NormalTok{) }\CommentTok{# coordinates of destination}
\KeywordTok{dist_google}\NormalTok{(}\DataTypeTok{from =} \NormalTok{o, }\DataTypeTok{to =} \NormalTok{d, }\DataTypeTok{mode =} \StringTok{"transit"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##   distances duration
## 1      2962     2118
\end{verbatim}

The code above takes an origin (\texttt{o}) and destination (\texttt{d})
and finds the travel time and distance. This is reported as 2,962 metres
and 2,118 seconds (35 minutes) respectively. To ensure that the result
was relevant to school travel, the times were calculated for arrival
time at 9am on a weekday. The function was repeated for all
origin-destination pairs. To ensure the correct OD data was being
generated, the results were plotted on an interactive map (Figure 2).
The results show the high variability of route distance versus Euclidean
distance (`circuity'), providing further evidence of inequalities in
public transport provision in the city (Figure 3).
\begin{figure}
\includegraphics[width=1\linewidth]{od-map} \caption{Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.}\label{fig:unnamed-chunk-4}
\end{figure}
\begin{figure}
\includegraphics[width=1\linewidth]{public-vs-euclid-dists} \caption{Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API. The colours represent the interquantile range of the Euclidean distances travelled to school.}\label{fig:euclid}
\end{figure}

\section{3. Area of study}\label{area-of-study}

The São Paulo Metropolitan Region (SPMR) is a large metropolis located
in the South West of Brazil. It extends for over 7,700 squared
kilometers and groups 38 different municipalities besides the
municipality of São Paulo itself. The estimated population in 2010 was
close to 19.5 million in 2010. The integrated transport system, to which
users can access through an electronic card (Bilhete Único) is composed
of a railway and bus network. The railway network comprises 78.4
kilometers in five subway lines and six suburban railway lines which
provide less frequent and slower service than the subway. Some of the
bus network operates on dedicated bus corridors, which amounted to 500
km in 2015. As can be seen in Figure 1, the richer areas of the SPMR are
better served by public transport, while an important number of lower
income areas is under-served.

\begin{figure}
\includegraphics[width=0.8\linewidth]{income_pc} \caption{Public transport system and income per capita in São Paulo, 2010.}\label{fig:inc-trans}
\end{figure}

Regarding the educational system, since the constitution of 1946,
primary education is compulsory. In 2005, the legislation changed to
increase the duration of the primary education, going from eight to nine
years. In 2010, all regions finished the transition. In 2009, a
constitutional amendment imposed compulsory education for the primary
and secondary levels. The constitution imposes that this resolution
should be fully implemented by 2016. By law, the state/municipalities
should guaranty the adequate provision (100\% coverage of the demand).
There are national and state educational funds that are distributed
among the schools according to the number of enrolled students and
existence of facilities (library, laboratories, etc.).

The structure of the educational system is composed by both public and
private schools. The public schools are divided in the following levels:
Pre school (\textit{clase de alfabetização}) -- 4-5 year-olds (municipal
administration); Fundamental (\textit{ensino fundamental}) comprises
nine years, and the age range is usually 6-14 year-olds (municipal
administration); Secondary (\textit{ensino médio}) lasts a minimum of
three years, and the age range is 15-18 year-olds (state
administration); and Post-secondary (\textit{ensino superior}) (state
and federal administration).

According to the School Census (\textit{Censo Escolar}) 2010\footnote{see
  \url{http://portal.inep.gov.br/basica-censo} for details on the School
  Census}, there are 10,251 schools in the SPMR, out of which 54\% are
public (22\% state / 32\% municipal) and 46\% are private.
Geographically, students are assigned to the closest school to their
place of residency or their parent's working place. For secondary
school, the student can ask for a vacancy in a different school after
providing a valid justification and under the condition of the
availability of vacancy.

The municipality is responsible for the organization of the pre-school
and fundamental level. Poorer municipalities in SPMR have, in general,
lower quality in terms of teachers and facilities. All the secondary
schools are administrated by the state, nevertheless, differences exists
depending of the location in the SPMR. In general, schools in peripheral
locations show lower quality scores.

As can be seen by comparing Figures 4 and 5, with Figure 1, students in
the SPMR disproportionally attend private secondary schools in higher
income areas, while the opposite is true for public secondary schools.
\begin{table}[]\centering \caption{Modal split by type of secondary school}
\begin{tabular}{lcc} \hline
\multicolumn{3}{c}{\textbf{Secondary Public School Students}}  \\  \hline
 & Percent & Duration (min)  \\ \hline
Public & 23 & 36.7 \\
Private & 9 & 15.8 \\
Active & 68 & 15.7 \\ \hline
\multicolumn{3}{c}{\textbf{Secondary Private School Students}}  \\  \hline
Public & 28 & 35.7 \\ 
Private & 47 & 17.3 \\
Active & 25 & 13.9 \\\hline
\multicolumn{3}{l}{\small Public transport includes bus, metro and railway.} \\
\multicolumn{3}{l}{\small Active transport includes walking and cycling.}  
\end{tabular}
\end{table}
\begin{figure}
\includegraphics[width=0.7\linewidth]{med_pu} \caption{The home location of students attending public secondary schools students in the SPMR.}\label{fig:unnamed-chunk-5}
\end{figure}\begin{figure}
\includegraphics[width=0.7\linewidth]{med_pr} \caption{The home location of students attending private secondary schools students in the SPMR.}\label{fig:unnamed-chunk-6}
\end{figure}

As Figure 6 shows, enrolment in public secondary schools is more
dispersed spatially, as in principle public schools are not clearly
geographically concentrated within the region. There is, however, an
important degree of variation in the school seats in each sub-area,
ranging from zero to 7,754.
\begin{figure}
\includegraphics[width=0.7\linewidth]{enrol} \caption{Enrollment in public secondary schools in the SPMR.}\label{fig:unnamed-chunk-7}
\end{figure}
Table 1 shows the model split by type of secondary school
student.\footnote{The source of this information is the 2007
  Origin-Destination Household Travel Survey (O-D Survey), carried out
  by the São Paulo Metropolitan Transport Authority Mêtro. We focus on
  trips made by secondary school students with an indicated motive
  ``education'' at the destination. The total number of trips (excluding
  missing values) was 11,845. Of these, 5.5 percent were multimodal. In
  those cases, we assign the mode of the trip leg with the largest
  duration. We add all trip leg durations (in minutes) to obtain the
  total trip duration.} Clearly, secondary school students attending
private schools disproportionally travel by car, compared to public
school students (9\% versus 47\%), even though higher income areas are
better served by public transport. This is in line with the findings of
S{á} et al. (2015). Around 67\% of public school students commute to
school by active travel (mostly walking, since the percentage of biking
is still relatively small), while 23\% commute by public transport,
which has much longer mean duration than other modes. Gomide (2003),
using data from the 1997 SPMR Origin-Destination survey, reports that
the motive ``education'' explained at least 60\% of trips in the SPMR of
individuals earning up to one minimum wage a month. This proportion
decreases as income increases, as other motives such as work and
leisure become more relevant.

Although the mean duration of travel to school remained stable over the
period 1997-2007 (S{á} et al. 2015), travel-to-school times could
increase as a result of several factors relevant to the SPMR, including
sub-urbanization of particular demographic groups, such as those with
higher income; increase in the size of schools leading to an increase in
school's catchment area; less than proportional increase in schools in
high population growth areas with poor connectivity; and institutional
changes towards more flexible school choice (Easton and Ferrari 2015).

\section{4. Method: measuring school
accessibility}\label{method-measuring-school-accessibility}

To define school accessibility, we start with the broad concept of
accessibility, as used in the transport literature. According to Geurs
and Van Wee (2004) (p.~128) accessibility is ``the extent to which
land-use and transport systems enable (groups of) individuals to reach
activities or destinations by means of a (combination of) transport
mode(s).'' Extended to our case, school accessibility then measures the
extent to which the existing built-environment and transport facilities
enable children and adolescents in school age to reach schools.

Ideally, a school accessibility measure would include information on the
home location of every student and the location of every school they
could potentially attend (as in Andersson, Malmberg, and {Ö}sth (2012)),
and some measure of commuting costs for any chosen mode (in terms of
time and money). Given that this level of detail is not available in our
data, we rely on area aggregates, implicitly assuming that the measure
at the centroid of the area applies homogeneously within the area.

Given these limitations, we first consider a cumulative-opportunity
measure for mode \(M\) (Boisjoly and El-Geneidy 2016) is defined as:

\[ CO_{i}^M= \sum_{j=1}^nO_{j}f(C_{ij}) \]

\[ f(C_{ij}) = \left\{ 
                \begin{array}{ll}
                  1\quad if\quad C_{ij}<=t\\
                  0\quad if\quad C_{ij}<t
                \end{array}
              \right.
\]

This measure counts the number of ``opportunities'' \(O\), in our case
schools available, from one area within a certain travel time threshold
by mode \(M\). \(C_{ij}\) is the travel cost (measured in time) between
the centroid of zone \(i\) and the centroid of zone \(j\), and
\(f(C_{ij})\) is a weight function.

This measure takes into account the spatial distribution of schooling
opportunities, but not the local demand for schooling. This is
particularly relevant for our analysis, since it could be the case that
some areas are disproportionally served with respect to the number of
potential students living within a certain travel distance. In order to
assess the mismatch between the demand and supply for schooling, we use
the sum of students in schools in each area (supply), and the sum of
individuals within the school grade age-group living in each area
(demand) in the following competitive accessibility measure (Shen 1998):

\[ CA_{i}^M= \sum_{j=1}^n\frac{O_{j}f(C_{ij})}{\sum_{k=1}^n(T_{k}f(C_{kj}))}\]

Where the numerator discounts the number of school seats in area \(j\),
\(O_{j}\), by how far area \(i\) is from area \(j\) using the same
function as before, and the denominator discounts the number of teens in
secondary school age living in zone \(k\), \(T_{k}\), by how far they
are from area \(j\). In other words, the numerator counts how many
schools seats can be reached from an area, while the denominator counts
how many teens can potentially reach the same area. In this way, the
discounted number of school seats places at each area is divided by the
potential students available to fill those places, and then summed in
order to obtain a single accessibility measure for each area.

We calculate both measures for commutes to school by public transport.
We set a threshold of 30 minutes based on the mean average travel times
by this mode (see Table 1). We assign a minimum time of zero for
commutes within the same area (i.e., schools within the same area are
added to the index in all cases).

\section{5. Results}\label{results}

Figure 7 shows the spatial distribution of the cumulative-opportunity
measure. The index has a mean value of 3,777, which indicates that for
teens residing in the average area, there are 3,777 school seats that
can be reached within a journey by public transport of up to 30 minutes.
The standard deviation is 3,485 indicates large variation in the level
of accessibility across areas.

As can be seen on the Figure 7, areas with low accessibility include not
only the expanded financial business district located on the central
south west area, where higher-income teens mostly attend private school
(see Figures 1 and 5), but also areas with deficient public transport
provision, such as some north and south eastern areas.

\begin{figure}
\includegraphics[width=0.7\linewidth]{cum_index} \caption{Cumulative opportunity index}\label{fig:unnamed-chunk-8}
\end{figure}

Figure 8 shows the spatial distribution of the competitive accessibility
measure. Although it largely follows the distribution of the cumulative
index, it corrects the value in some areas based on the place of
residence of students across the SPRM. To interpret the value of the
index, recall that if all teens in an area attended secondary schools in
public secondary schools in their area (i.e., no student would attend
private school or travel outside her own area to attend school), the
index would take the value of 1. Then, in areas with a value lower than
one, the provision of public schools and public transport is such that
there is local under-provision, in the sense that the number of schools
seats that can be reached from the area is smaller than the number of
teens that can potentially reach the area (the opposite is true for
values higher than one). Note that some relatively close by areas,
especially in the eastern part of the city, have striking differences in
the level of accessibility. In some cases, an area can have twice the
level of accessibility than the most proximate area. This can be due to
differences in the local provision of public schools, public transport
or the number of students living within the reach of the areas.
\begin{figure}
\includegraphics[width=0.7\linewidth]{comp_index} \caption{Competitive accessibility index}\label{fig:unnamed-chunk-9}
\end{figure}
To understand the effects of changes in public school provision on
school accessibility, we conduct a comparative static analysis using our
competitive index. We compare the changes in school-accessibility by
accessibility quantiles before and after the implementation of a policy
that reallocates students attending school in areas with lower than
average provision to schools in areas with higher than average
provision. The experiment mimics a policy that aims at increasing scale
economies in the provision of public education by centralizing provision
in larger educational establishments, which is the rationale behind the
policy proposal made by the São Paulo state government (Ortellado 2015).

To implement the policy experiment, we assign students in areas with
lower than average enrolment to the next area with higher than average
enrolment, while keeping the distribution of teens and public transport
constant. In other words, total enrolment is the same before and after
the change, but the distribution of public secondary schools increases
in areas where it was already higher than average, and disappears in
areas where it was lower than average. The change is equivalent to
closing down 1,691 schools, or reallocating 163,231 students. Although
seemingly extreme, it is worth noting again that the proposal of the São
Paulo state government was estimated to impact 300,000 secondary
students attending public school (Ortellado 2015). We then compare the
changes in school accessibility for different quartiles before and after
this change. Table 2 summarizes the results.

\begin{longtable}[c]{@{}lrrr@{}}
\caption{Mean accessibility before and after
redistribution}\tabularnewline
\toprule
Quantile & Pre-change & Post-change & Change (\%)\tabularnewline
\midrule
\endfirsthead
\toprule
Quantile & Pre-change & Post-change & Change (\%)\tabularnewline
\midrule
\endhead
1st & 0.446 & 0.350 & -21.566\tabularnewline
2nd & 0.844 & 0.822 & -2.592\tabularnewline
3rd & 1.843 & 2.124 & 15.213\tabularnewline
4th & 4.280 & 5.483 & 28.099\tabularnewline
\bottomrule
\end{longtable}

\newpage

The policy reduces further the accessibility level of areas in the first
and second quantiles of accessibility. Areas in the first quantile
experience an average reduction of 21\% in school accessibility levels
as a result of the changes in the spatial distribution of schools. In
the mean time, areas that had high levels of accessibility to start
with, experience an increase. The results suggest that concentrating the
provision of public secondary schools accentuates the already large
differences in accessibility levels across areas.

\section{6. Conclusion and discussion}\label{conclusion-and-discussion}
